---
title: "Analyses and Background Information in support of the manucript: Forest structural complexity and biomass predict first-year carbon cycling responses to disturbance, Gough et al. "
author: "Gough et al."
date: "4/30/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(kableExtra)
#source("./R/utilities.R")
```

## Overview

This document accompanies the manuscript, "Forest structural complexity and biomass predict first-year carbon cycling responses to disturbance" by Gough et al. submitted to Ecosystems in 2020. This analysis focuses on first year carbon cycling responses from the Forest Resilience and Threshold Experiment (FoRTE): <https://fortexperiment.github.io/fortedata/>.

Additional analyses scripts written for SAS are availble in this download.

## Forest structural complexity and biomass predict first-year carbon cycling responses to disturbance

Authors:  Christopher M. Gough^1^, Jeff W. Atkins^1^, Ben Bond-Lamberty^2^, Elizabeth A. Agee^3^, Kalyn R. Dorheim^2^ Robert T. Fahey^4^, Maxim S. Grigri^1^, Lisa T. Haber^1^, Kayla C. Mathes^1^, Stephanie C. Pennington^2^, Alexey N. Shiklomanov^5^, Jason M. Tallant^6^ 

1. Department of Biology, Virginia Commonwealth University, Box 842012, 1000 West Cary St., Richmond, VA 23284
2. Joint Global Change Research Institute, Pacific Northwest National Laboratory, 5825 University Research Ct, College Park, MD 20740
3. Environmental Sciences Division and Climate Change Science Institute, Oak Ridge National Laboratory, Oak Ridge, TN, 37831, USA
4. Department of Natural Resources and the Environment & Center for Environmental Sciences and Engineering, University of Connecticut, 1376 Storrs Road, Storrs, CT 06269
5. NASA Goddard Space Flight Center, Greenbelt, MD 20771, USA
6. University of Michigan, Biological Station and Department of Ecology and Evolutionary Biology, Pellston, MI 49769, USA

*Corresponding Author: <cmgough@vcu.edu>
Submitted to: Ecosystems <https://www.springer.com/journal/10021>


### Abstract

The pre-disturbance vegetation characteristics that confer initial carbon (C) cycling resistance to disturbance are not well known. To address this gap, we initiated the Forest Resilience Threshold Experiment (FoRTE), a manipulative study comprised of replicated factorial combinations of four levels (control, 45%, 65% and 85% gross defoliation) of disturbance severity and two disturbance types (targeting upper or lower canopy strata) implemented by stem girdling >3600 trees. Applying a standardized stability framework in which resistance was calculated as the first-year natural log response ratio of disturbance and control treatments, we investigated to what extent pre-disturbance levels of species diversity, aboveground woody biomass, leaf area index, and canopy rugosity – a measure of structural complexity – predict the initial resistance of subcanopy light-saturated leaf CO~2~ assimilation (A~sat~), aboveground wood NPP (ANPP~w~) and soil respiration (R~s~) to phloem-disrupting disturbance. We found that above-ground C cycling processes, Asat and ANPPw, were highly resistant to increases in disturbance severity, while Rs resistance declined as severity increased. Disturbance type had no effect on first-year resistance. Pre-disturbance aboveground woody biomass was a positive predictor of ANPP~w~ resistance and, along with canopy rugosity, was negatively related to R~s~ resistance. Subcanopy Asat resistance was not related to pre-disturbance vegetation characteristics. Stability of C uptake processes along with Rs declines suggest the net C sink was sustained in the initial months following disturbance. We conclude that biomass and complexity are significant, but not universal, predictors of initial C cycling resistance to disturbance. Our findings highlight the utility of standardized stability measures when comparing functional responses to disturbance.


```{r data.import}
df <- read.csv("./data/forte_resistance_data_yr_one.csv")

# change severity from integer to factor
df$severity <-as.factor(df$severity)
```

### Question One: Do key carbon cycling processes respond similarly to disturbance?

```{r question.one}
df %>%
  select(rep, severity, type, rNPP, rRs, rAmax) %>%
  pivot_longer(cols = c(rNPP, rRs, rAmax),
               values_to = "resistance") -> df.resist


# GLM model
# original SAS code:
# proc sort; by funct; run;
# proc glm;
# class rep sev trt funct;
# model R = rep sev trt;
# means sev/lsd alpha = 0.1; by funct;
# run;

resist.model <- lm(resistance ~ rep + severity + type, data = df.resist)
summary(resist.model)

```
## Q2: Does carbon cycling process resistance change as a function of disturbance severity and canopy strata affected?

## Q3: Do pre-disturbance canopy structural complexity and LAI affect carbon cycling process resistances?
```{r samp.var}

# Let's look at means of variables
df %>%
  select(-type) %>%
  group_by(rep, severity) %>%
  dplyr::summarize_all(list(~mean(.), ~var(.))) -> df.means.var

knitr::kable(df.means.var[,c(1:2, 3:10)], caption = "Variable means for year 2019") %>%
  kable_styling() 

knitr::kable(df.means.var[,c(1:2, 11:18)], caption = "Variable variance for year 2019") %>%
  kable_styling() 
  
```

Next...Testing LAI and resistance

```{r rugosity}
# sas code:  
# proc glm data=resist2;
# class severity;
# model rNPP rRs rAmax = rugosity rugosity*severity;
# run;

# R version of the SAS GLM 

# NPP resistance
lm.rNPP <- lm(rNPP ~ rugosity + rugosity*severity, data = df)
summary(lm.rNPP)

# Soil Respiration (Rs) resistance
lm.rRs <- lm(rRs ~ rugosity + rugosity*severity, data = df)
summary(lm.rRs)

# Amax resistance
lm.rAmax <- lm(rAmax ~ rugosity + rugosity*severity, data = df)
summary(lm.rAmax)
```

Next...Testing structural (rugosity) and resistance

```{r LAI}
# sas code:  
# proc glm;
# class severity;
# model rNPP rRs rAmax = LAI LAI*severity;
# run;


# R version of the SAS GLM 

# NPP resistance
lm.rNPP.LAI <- lm(rNPP ~ LAI + LAI*severity, data = df)
summary(lm.rNPP.LAI)

# Soil Respiration (Rs) resistance
lm.rRs.LAI <- lm(rRs ~ LAI + LAI*severity, data = df)
summary(lm.rRs.LAI)

# Amax resistance
lm.rAmax.LAI <- lm(rAmax ~ LAI + LAI*severity, data = df)
summary(lm.rAmax.LAI)
```



```{r biomass}
# sas code:  
# proc glm;
# class severity;
# model rNPP rRs rAmax = biomass biomass*severity;
# run;


# R version of the SAS GLM 

# NPP resistance
lm.rNPP.biomass <- lm(rNPP ~ biomass + biomass*severity, data = df)
summary(lm.rNPP.biomass)

# Soil Respiration (Rs) resistance
lm.rRs.biomass <- lm(rRs ~ biomass + biomass*severity, data = df)
summary(lm.rRs.biomass)

# Amax resistance
lm.rAmax.biomass <- lm(rAmax ~ biomass + biomass*severity, data = df)
summary(lm.rAmax.biomass)
```


```{r shannon}
# sas code:  
# proc glm;
# class severity;
# model rNPP rRs rAmax = shannon shannon*severity;
# run;


message("Testing")
# R version of the SAS GLM 

# NPP resistance
lm.rNPP.shannon <- lm(rNPP ~ shannon + shannon*severity, data = df)
summary(lm.rNPP.shannon)

# Soil Respiration (Rs) resistance
lm.rRs.shannon <- lm(rRs ~ shannon + shannon*severity, data = df)
summary(lm.rRs.shannon)

# Amax resistance
lm.rAmax.shannon <- lm(rAmax ~ shannon + shannon*severity, data = df)
summary(lm.rAmax.shannon)
```

